Unifying the communicable disease spreading paradigm with Gompertzian growth

### 2022-07-14 ReviewerResponse.md

Reviewer responses
------------------

Reviewer 1:
-----------
The reviewer makes the following conjecture:
"The fact that the COVID-19 mortality can be accurately reproduced by the Gompertz curve does not seems to be very surprising taking into account that this dynamical dependence was initially developed to describe mortality curve and confirmed in this application by a variety of examples during almost two centuries"

While Gompertzian growth has been shown to exist in natural growth phenomena like cancer growth, plant leaf growth, etc., it has not previously been shown in the communicable disease paradigm before covid, e.g. at large spatial distances (e.g. across countries). This should indeed be surprising for the following reason: In smaller spatial scales, e.g. bacterial growth as the reviewer alludes to in Lavrova et al., the exponentially limiting growth factor is quite plausible. In the larger national scales, this exponential limiting factor is not plausible nor possible under the SIR/SIRD model, as we show in the introduction. The SIR/SIRD models imply a logistic growth curve, which is indeed not what we see. This conclusion is only possible due to our daily cadence of mortality data, not available for previous epidemics. In a weekly cadence, for example, the separation between the Gompertz and the Logistic is more tenuous and less convincing.

As we discuss in the paper and as has been discussed by references like Petroni et al., Gompertzian growth implies coherent or collaborating growth. In a communicable disease paradigm, we show that this collaboration can only result from infinite interactions and communication speed as exemplified in our proposed network in section 3.1. This is due to the constraint that people must be in contact to propagate the disease. In the reviewer's example from Lavrova, this constraint is not present and is indeed modeled as a ubiquitous oxygen concentration. This is a crucial difference that seems to have been missed by the reviewer.

Further regarding the reviewers comment:
"It also should be noted that the statement "a discrepancy between reality and the current theory of communicable diseases" does not seems to be correct because the classic (either SIR or SEIR)  theory of communicable diseases does not distinguish between "Removed" and "Recovered" persons as the variable "R". [...] Thus, a more motivated statement should be about developing a subdivision of an epidemiological model aimed at distinguishing between two outcomes."

This is exactly why we include a model that does distinguish between "Removed" and "Recovered" in the SIRD model and show that it leads to the logistic curve. Further, the premise of the paper is not to see what additional parameters must be added to the SIRD-like models to obtain Gompertzian growth, but rather to show that with only 2 parameters, under the environmental stressor model, the observations are accurately predicted. We do in fact leave it as an open question when we address potential networks that can lead to Gompertz, but do provide one such network in section 3. This seems much more "motivated" due to the parsimonious nature of the environmental stressor model, as opposed to adding to a model that already has more parameters and yet cannot accurately predict observations.

Regarding comment:
"Although the authors introduced SIRD model further in the text, the clear description of interplay about assumptions of both (SIR and SIRD) approaches should be discussed with the premises of both."

As we already discuss the SIRD and the SIR model based on the same premises, this comment is not clear. We introduce the "Dead" or "Removed" group for the exact reasons outlined by the reviewer himself above. A further elaboration of what kind of "premises" should be discussed is unclear.

"In addition, it should be pointed out that Ref. [5] already references the work Anastassopoulou at al. (2020). Data-based analysis, modelling and forecasting of the COVID-19 outbreak. PloS One, 15, e0230405 as that where SIRD model was introduced." 
It is not clear how this is relevant to the manuscript. We chose to follow Carletti due to the clarity of their exposition, not because they were the first ones to use the SIRD model.

Our modeling framework of the Gompertz does of course have wide implications with regards to the mechanisms of growth: We postulate that growth is taking place with instant communication between units. In the case of plant photosynthesis this is not so far fetched considering that electrons in the electron transport chain are already shown to be non-local, i.e. wavelike rather than particle like.  See for example G. Panitchayangkoon et al. "Long-lived quantum coherence in photosynthetic complexes at physiological temperature" PNAS 2010. This of course breaks down the chemical paradigm where molecules are localized in regular biological processes.

Although it was not the wish of the authors to present too much observational data that does not fit with the current virus-theory of covid, but rather focus on a mathematical model that does indeed explain this simple set of daily mortality data with a simple 2-parameter model, it may be illuminating to the reviewers and editor to note that many animals have been observed to die at anomalous rates during the spring of 2020.

- Fischer L, Peters M, Merbach S, Eydner M, Kuczka A, Lambertz J, Kum- merfeld M, Kahnt K, Weiss A, Petersen H. Increased mortality in wild tits in north rhine-westphalia (germany) in 2020 with a special focus on suttonella ornithocola and other infectious pathogens. European Journal of Wildlife Research 2021;67(3). doi:10.1007/s10344-021-01500-7.
- Peris ́e-Barrios AJ, Tomeo-Mart ́ın BD, G ́omez-Ochoa P, Delgado-Bonet P, Plaza P, Palau-Concejo P, Gonz ́alez J, Ortiz-D ́ıez G, Mel ́endez- Lazo A, Gentil M, Garc ́ıa-Castro J, Barbero-Fern ́andez A. Humoral responses to SARS-CoV-2 by healthy and sick dogs during the COVID- 19 pandemic in spain. Veterinary Research 2021;52(1). doi:10.1186/ s13567-021-00897-y. 
- Duff P, Fenemore C, Holmes P, Hopkins B, Jones J, He M, Everest D, Rocchi M. Disease surveillance in england and wales, december 2019. Vet- erinary Record 2020;186(1):15–6. doi:10.1136/vr.m68
- HuB,WeiH,FanZ,SongY,ChenM,QiuR,ZhuW,XuW,XueJ,Wang F. Emergence of rabbit haemorrhagic disease virus 2 in china in 2020. Veterinary Medicine and Science 2020;7(1):236–9. doi:10.1002/vms3.332.
- Fukui H, Shimoda H, Kadekaru S, Henmi C, Une Y. Rabbit hemorrhagic disease virus type 2 epidemic in a rabbit colony in japan. Journal of Vet- erinary Medical Science 2021;83(5):841–5. doi:10.1292/jvms.21-0007.
- van Aarde RJ, Pimm SL, Guldemond R, Huang R, Mar ́e C. The 2020 elephant die-off in botswana. PeerJ 2021;9:e10686. doi:10.7717/peerj. 10686.
- Kambayashi Y, Bannai H, Tsujimura K, Hirama A, Ohta M, Nemoto M. Outbreak of equine coronavirus infection among riding horses in tokyo, japan. Comparative Immunology, Microbiology and Infectious Diseases 2021;77:101668. doi:10.1016/j.cimid.2021.101668.
- King S, Rajko-Nenow P, Ashby M, Frost L, Carpenter S, Batten C. Out- break of african horse sickness in thailand, 2020. Transboundary and Emerg- ing Diseases 2020;doi:10.1111/tbed.13701.
- Lu G, Pan J, Ou J, Shao R, Hu X, Wang C, Li S. African horse sickness: Its emergence in thailand and potential threat to other asian countries. Transboundary and Emerging Diseases 2020;doi:10.1111/tbed.13625.
- Castillo-Olivares J. African horse sickness in thailand: Challenges of controlling an outbreak by vaccination. Equine Veterinary Journal 2020;53(1):9–14. doi:10.1111/evj.13353.

Taken as a whole, this cannot be explained by a virus. Indeed it would require the "virus" to spread to birds across the world simultaneously which seems quite implausible. A more plausible mechanism is an environmental stressor caused by a planetary change in the environment.

Minor text issues:
 (page 8): the sentence "where the bra-ket notation" could be misleading because it induces associations with quantum mechanical terms, inner products, etc. Here angle brackets denote simply standard statistical averaging.
 Noted, this will be changed to a simple expectation symbol "E[.]"

(page 9): what does it mean "planetary perturbation"?
This is the postulated cause of the environmental stressor. We can elaborate on this in the manuscript.

Reviewer 2:
-----------
This review is difficult to take seriously as it seems the reviewer has not understood the concepts introduced in the paper, as exemplified by
1. A lack of a reasoned discussion of why the paper should not be published.
2. The use of words like "esoteric". This word implied a something "mysterious" that cannot be explained, even though thorough explanations and even examples in section 3 are given.
3. The use of the word "kind of" implies a lack of understanding.
4. The reference given as Rodrigues et al. does not address the reason why Gompertz is observed, but merely shows that the observations can be fit with such a model, as has been cited by multiple papers in the manuscript. This reference in particular is a far cry from explaining the mechanistic underpinnings of the Gompertz model including its unification of the communicable disease paradigm based on a network of infected and susceptible people, thus obviating the statement "The present manuscript does not have an additional value." which also seems to confuse the reviewers value system with that of the general scientific community.
5. No discussion of the technical aspects of the manuscript.

Taken together it is clear that reviewer 2 did not engage with the material at the sufficient level to understand it.

To the Editor
-------------
I thank you for considering this manuscript for publication, but I must say that the quality of the reviews was not what I expected. This paper has the potential to have large ramifications in the scientific community as it not only postulates an alternative interpretation of observations with regards to covid, but also offers an alternative theory for Gompertzian growth in general and will thus affect our understanding of cancer growth, plant growth, etc.

I ask if you could suggest this paper to a few other reviewers that will hopefully engage more authentically with the material. Especially reviewer 2 does not seem to have the level of competence for this kind of job. Reviewer 1 (which I suspect is Postnikov) has some reasoned points, but are based on a misunderstanding in my view, and thus render me without tangible corrections I can make to improve the delivery of the manuscript's message.

### 2022-07-14 elsarticle/elsarticle.tex

This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves. What is causing such a discrepancy between reality and the current theory of communicable diseases \citep{castro2020turning}?

One could approach this conundrum from parametrization of the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}. How would $\phi(t)$ have to behave? Start by recalling the SIR model for a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
\begin{align}
\label{eq:SIR}
\frac{dS}{dt}& = -\beta \frac{IS}{N}\nonumber\\
\frac{dI}{dt}& = \beta \frac{IS}{N} - \alpha I.\\
\frac{dR}{dt}& = \alpha I\nonumber,
\end{align}
omitting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.

@@

 At this limit the relative growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
 The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning. 
 One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
 Rypdal and Rypdal \cite{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries. In fact, Wang et al. \cite{wang2012richards} showed that values of $\theta<1$ are not biologically realistic under a similar SIR model as the one given above.

 Perhaps a more likely scenario from which a Gompertz curve would emerge is the selective infection of central nodes in the transmission network resulting in an immediate decrease in relative growth. 
 On the other hand, an infection of peripheral nodes should cause immediate exponential growth. 
 Herrmann and Schwartz \cite{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve. 
 Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. 
 We will touch upon how the Gompertz curve emerges from one such network below, under some caveats.

\subsection{The SIR model family is almost Gompertzian}
Without the linearization and the somewhat arbitrary $\theta$-parametrization, one can still obtain Gompertzian growth from the communicable disease models, i.e. a straight line under a double-log transform of Eq. \ref{eq:gomp_solution}, viz.
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this, we follow Carletti et al. \cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This requires an addition to our original set of equations \ref{eq:SIR}, with an extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

Algebraic manipulations reveal that the diseased group is described by a single equation, viz.
\begin{equation}
\label{eq:SIRD_D}
\frac{dD}{dt} = \eta( 1 - e^{-\xi D}) - \kappa D,
\end{equation}
where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is again synonymous to the initial susceptible pool of people. This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery rates, $\delta$ and $\alpha$. 
However, when only modeling the diseased, knowledge of $N$ is irrelevant and the three parameters in Eq. \ref{eq:SIRD_D} are sufficient.

Note here that small values of D follow logistic growth, viz.
\begin{equation}
\frac{dD}{dt} = c D (1 - D/K),
\end{equation}
obtained with a second order expansion of the exponential term in Eq. \ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$. The SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.

With the full 3-parameter single ODE of the diseased group in the SIRD model, one can obtain a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity (Fig. \ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in Eq. \ref{eq:GompertzODE}, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality,
\begin{equation}
\label{eq:GompertzODE2param}
\frac{d}{dt}\ln{Y(t)} = -\beta\left[\ln{Y(t)} - \ln {Y_{\infty}}\right].
\end{equation}
Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters \citep{roda2020difficult}.

Note also that the SIRD model above considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond} which does not consider network effects explicitly. Rather, all entities are connected and communicating instantaneously as shown by Mombach et al. \cite{mombach2002mean}.
However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain. 
We are thus prompted to look for another model which can explain the observed mortality patterns.

@@

\section{An alternative theory for observed mortality}
\label{seq:alt}
An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability. 
This could explain the immediate dampening of mortality growth observed not explainable by a SIR model. 
It could also explain the lack of correlation between population density and mortality (or infection) rates observed by some \citep{Hamidi2020,Hamidi2020a,Carozzi2020,Arpino2020,khavarian2021high,barak2021urban}, in addition to the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021,NINA2020}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.

>>>>

This phenomenon is recorded by others \citep{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}, showing Gompertz curves at national levels instead of the traditionally predicted logistic curves, and are made possible in part due to the increased cadence and quality of data acquisition compared with earlier epidemics. What is causing such a discrepancy between reality and the current theory of communicable diseases \citep{castro2020turning}?

We start by reviewing a basic infectious disease model, namely the Susceptible-Infected-Recovered (SIR) model under a time-dependent infection/recovery rate ratio, $\phi(t)$ \citep{kermack1927contribution}, and show under what conditions the Gompertz curve could emerge. First assume a pool of susceptible people of size $N$ evolving between the three states: susceptible, $S(t)$, recovered, $R(t)$ and infected, $I(t)$,
\begin{align}
\label{eq:SIR}
\frac{dS}{dt}& = -\beta \frac{IS}{N}\nonumber\\
\frac{dI}{dt}& = \beta \frac{IS}{N} - \alpha I.\\
\frac{dR}{dt}& = \alpha I\nonumber,
\end{align}
omitting the argument $t$ in each variable for brevity and where $\alpha$ and $\beta$ in this context signify recovery and infection rates.

@@

 At this limit the relative growth rate approaches infinity as $Y \rightarrow 0$, which seems odd under the hypothesis that the pathogen has just started spreading. 
 The Gompertzian limit also implies a decreasing relative growth rate from the very first time point, which under the SIR model seems unlikely given the large pool of susceptible people in the beginning. 
 One would rather expect a near-constant relative growth rate in the beginning due to a disease propagation stage. 
 Rypdal and Rypdal \cite{Rypdal2020} suggest that the decreasing relative growth rate is caused by social and political mitigating efforts, but these hardly justify such coherent and consistent mortality characteristics across countries, particularly with an exponentially decreasing growth rate. In fact, Wang et al. \cite{wang2012richards} showed that any values of $\theta<1$ are not physically realistic under a similar SIR model as the one given above.

 Perhaps a more likely scenario from which a Gompertz curve would emerge is the selective infection of central nodes in the transmission network resulting in an immediate decrease in relative growth. 
 Herrmann and Schwartz \cite{herrmann2020covid} studied a networked SIR model on a variety of networks, but did not elaborate on a possible fit to a Gompertz curve. Estrada and Bartesaghi \cite{estrada2022networked} found that the Gompertz can emerge in later stages of an SIS (Susceptible-Infected-Susceptible) model, but not in the beginning of an SIR-like model.
 Although it may be possible to realize Gompertzian growth from a special network, firm theoretical work has yet to be done to show this connection. 
 We will touch upon how the Gompertz curve emerges from one such network below, under some caveats.

\subsection{The extended SIR model family is almost Gompertzian}
Without the linearization and the $\theta$-parametrization of the Richard's model, one can still almost obtain Gompertzian growth from the communicable disease models. Under a double-log transform of Eq. \ref{eq:gomp_solution}, the growth curve is a simple straight line which can be seen as a tell-tale sign of a Gompertz curve,
\begin{equation}
\label{eq:GOMP_D}
\ln{(\ln{(Y_{\infty}/Y)})} = -\beta t + k,
\end{equation}
with $k=\ln{\ln{\frac{Y_{\infty}}{Y_{0}}}}$. To show this, we follow Carletti et al. \cite{carletti2020covid} and consider the extended version of the SIR the model by including a group of diseased, $D(t)$, such that $N = S(t) + I(t) + R(t) + D(t)$, also called the SIRD model. This requires an addition to our original set of equations \ref{eq:SIR}, with an extra equation for the diseased group's growth at some rate $\delta$ relative to the current infected group,
\begin{equation}
\frac{dD}{dt} = \delta I.
\end{equation}

Algebraic manipulations reveal that the diseased group is described by a single equation, viz.
\begin{equation}
\label{eq:SIRD_D}
\frac{dD}{dt} = \eta( 1 - e^{-\xi D}) - \kappa D,
\end{equation}
where $\eta = \delta N$, $\xi = \beta/(\delta N)$, $\kappa = \delta + \alpha$, and where $N$ is again synonymous to the initial susceptible pool of people. This pool of people cannot be obtained from deaths alone, but can be inferred by assuming a known ratio between mortality and recovery rates, $\delta$ and $\alpha$. 
However, when only modeling the diseased as we do here, knowledge of $N$ is irrelevant and the three parameters in Eq. \ref{eq:SIRD_D} are sufficient.

In the beginning of an epidemic one can assume small values of D and thus the SIRD model follows logistic growth with
\begin{equation}
\frac{dD}{dt} = c D (1 - D/K),
\end{equation}
obtained with a second order expansion of the exponential term in Eq. \ref{eq:SIRD_D} and with $c=\eta\xi - \kappa$ and $K=2c/\eta\xi^2$. Both the SIR and SEIR model exhibits similar properties and their discrepancy with initial observations related to the Coronavirus epidemic has been noted by others \citep{vattay2020forecasting}.

Without this approximation, the full 3-parameters in Eq. \ref{eq:SIRD_D} of the diseased group can show a curve quite close to a straight line in the double-log domain even in the initial observations, although there will always be a non-zero concavity (Fig. \ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in Eq. \ref{eq:GompertzODE}, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality,
\begin{equation}
\label{eq:GompertzODE2param}
\frac{d}{dt}\ln{Y(t)} = -\beta\left[\ln{Y(t)} - \ln {Y_{\infty}}\right].
\end{equation}
Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters \citep{roda2020difficult}.

Note also that the SIRD model above considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \citep{smilkov2014beyond} which does not consider network effects explicitly. Rather, all entities are connected and communicating instantaneously as shown by Mombach et al. \cite{mombach2002mean}.
However, even with such strong assumptions, it is odd that observed mortality never exhibits a stage of near-exponential growth as this macroscopic SIRD model predicts, but rather a constant negative slope in the double-log domain.

Other more general observations that conflict with these communicable disease models is 1) the lack of correlation between population density and mortality (or infection) rates \citep{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between intergenerational relationships and fatality rates \citep{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \citep{Fischer2021}, dogs \citep{PeriseBarrios2021}, rabbits \citep{Duff2020,Hu2020,Fukui2021}, elephants \citep{Aarde2021}, and horses \citep{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}.
We are thus prompted to look for another model which can explain the observed mortality patterns.

@@

\section{An alternative theory for observed mortality}
\label{seq:alt}
An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an external stressor, initiating a stress response to eventually bring mortality rates back to stability.
